On Adding Pattern Matching to Haskell-based Deeply Embedded Domain Specific Languages

### 2020-10-08 paper/paper.tex
Comment out Church encoding and C backend sections

\subsection{Connection to Church encodings}
\label{sec:Church}
% \begin{todo}
%   Work on this subsection. Maybe this shouldn't be a separate section at all and
%   maybe its contents should be incorporated throughout the rest of the paper.
% \end{todo}

Algebraic datatypes can be represented as lambda calculus terms using
\textit{Church encodings}.~\cite{Boehm:1985:AutoSynth} For example, the pair constructor can be
represented as

\begin{lstlisting}
church_mkPair
  :: a -> b -> (a -> b -> r) -> r
church_mkPair x y = \ f -> f x y
\end{lstlisting}

Coproducts can be represented with the constructors

\begin{lstlisting}
church_mkLeft
  :: a -> (a -> r) -> (b -> r) -> r
church_mkLeft x = \ f g -> f x

church_mkRight
  :: b -> (a -> r) -> (b -> r) -> r
church_mkRight y = \ f g -> g y
\end{lstlisting}

In this representation:

\begin{itemize}
  \item A pair of type \ttt{A} and type \ttt{B} is represented with a value of type
\ttt{forall r. (A -> B -> r) -> r}.

  \item A coproduct of type \ttt{A} and type \ttt{B} is represented with a value of type
\ttt{forall r. (A -> r) -> (B -> r) -> r}.
\end{itemize}

Both of those representations represent the type as a function that performs the
corresponding case analysis. A deep embedding of the operations
\ttt{churchMkPair}, \ttt{church\_mkLeft} and \ttt{church\_mkRight} is given
by the type

\begin{lstlisting}
data Church r where
  ChurchPair
    :: a -> b -> (a -> b -> r) ->
       Church r
  ChurchLeft
    :: a -> (a -> r) -> (b -> r) ->
       Church r
  ChurchRight
    :: b -> (a -> r) -> (b -> r) ->
       Church r
\end{lstlisting}

with the evaluation function

\begin{lstlisting}
churchEval :: Church r -> r
churchEval (ChurchPair x y f) = f x y
churchEval (ChurchLeft x f g) = f x
churchEval (ChurchRight y f g) = g y
\end{lstlisting}

However, we only want to keep track of the functions used for elimination. By moving
the values actually being contained in the products and coproducts out to the
evaluation function, we arrive at

\begin{lstlisting}
data Church' t r where
  ChurchPair'
    :: (a -> b -> r) ->
       Church' (a, b) r

  ChurchEither'
    :: (a -> r) -> (b -> r) ->
       Church' (Either a b) r

churchEval' :: Church' t r -> t -> r
churchEval' (ChurchPair' f) (x, y) = f x y
churchEval' (ChurchEither' f g) e =
  case e of
    Left  x -> f x
    Right y -> g y
\end{lstlisting}

The \verb|Church'| type forms a large part of the foundation of the pattern
matching aspect of the \verb|E| type. A rough correspondence between the two
is given in the table below.\\

{\small
\noindent\begin{tabular}{|p{3.4cm}|p{4.4cm}|}
  \hline
  \ttt{Church' (a, b) r} & \ttt{E (SumMatch (a, b) r} \\
  \hline
  \ttt{Church' (Either a b) r} & \ttt{E (ProdMatch (Either a b) r)}\\
  \hline
\end{tabular}
}

@@

% \newpage
\section{C Backend}
\label{sec:CBackend}

\begin{todont}
  Should this section be expanded?
\end{todont}

The C backend makes use of a small EDSL for generating C code. See
Appendix B for implementation details of this ESDL.

The following \ttt{struct}s provide the C representation for Haskell expressions:

\begin{lstlisting}
typedef enum var_type_tag {
  EXPR_INT
, EXPR_PAIR
...
, EXPR_UNIT
} var_type_tag;

typedef enum semantic_type_tag {
  NO_SEMANTIC_TAG
, EXPR_COMPLEX     // Complex numbers
} semantic_type_tag;

struct closure_t;

typedef struct var_t {
  var_type_tag tag;
  semantic_type_tag semantic_tag;
  void* value;
} var_t;

typedef struct closure_t {
  var_t* fv_env;
  var_t (*fn)(var_t, struct closure_t*);
} closure_t;
\end{lstlisting}

The \ttt{semantic\_tag} allows type information preserved by the \ttt{Typeable} constraints
in the EDSL to be brought over to the generated C code. In our implementation, this is
used to distinguish between pairs of \ttt{Double}s and \ttt{Complex Double} values. In the
case of a \ttt{Complex Double}, the \ttt{var\_t} would have a \ttt{tag} field with the value
\ttt{EXPR\_PAIR} and a \ttt{semantic\_tag} field with the value \ttt{EXPR\_COMPLEX}.

A C code generation function \ttt{genC} is implemented:

\begin{lstlisting}
genC :: E a -> String
...
\end{lstlisting}

To use this, for example, with the previously mentioned \ttt{isEven} implementation we
can write:
% \newpage
\begin{lstlisting}
isEvenE :: E (Int -> Bool)
isEvenE =
  externalize (
    case x == 0 of
      True -> True
      False ->
        case x == 1 of
          True -> False
          False -> isEvenE (x - 2))

isEvenC :: String
isEvenC = genC isEvenE
\end{lstlisting}

@@

  \begin{wrapfigure}{R}{1.2\linewidth}

@@

\resizebox{1\linewidth}{!}{

>>>>

% \subsection{Connection to Church encodings}
% \label{sec:Church}
% % \begin{todo}
% %   Work on this subsection. Maybe this shouldn't be a separate section at all and
% %   maybe its contents should be incorporated throughout the rest of the paper.
% % \end{todo}

% Algebraic datatypes can be represented as lambda calculus terms using
% \textit{Church encodings}.~\cite{Boehm:1985:AutoSynth} For example, the pair constructor can be
% represented as

% \begin{lstlisting}
% church_mkPair
%   :: a -> b -> (a -> b -> r) -> r
% church_mkPair x y = \ f -> f x y
% \end{lstlisting}

% Coproducts can be represented with the constructors

% \begin{lstlisting}
% church_mkLeft
%   :: a -> (a -> r) -> (b -> r) -> r
% church_mkLeft x = \ f g -> f x

% church_mkRight
%   :: b -> (a -> r) -> (b -> r) -> r
% church_mkRight y = \ f g -> g y
% \end{lstlisting}

% In this representation:

% \begin{itemize}
%   \item A pair of type \ttt{A} and type \ttt{B} is represented with a value of type
% \ttt{forall r. (A -> B -> r) -> r}.

%   \item A coproduct of type \ttt{A} and type \ttt{B} is represented with a value of type
% \ttt{forall r. (A -> r) -> (B -> r) -> r}.
% \end{itemize}

% Both of those representations represent the type as a function that performs the
% corresponding case analysis. A deep embedding of the operations
% \ttt{churchMkPair}, \ttt{church\_mkLeft} and \ttt{church\_mkRight} is given
% by the type

% \begin{lstlisting}
% data Church r where
%   ChurchPair
%     :: a -> b -> (a -> b -> r) ->
%        Church r
%   ChurchLeft
%     :: a -> (a -> r) -> (b -> r) ->
%        Church r
%   ChurchRight
%     :: b -> (a -> r) -> (b -> r) ->
%        Church r
% \end{lstlisting}

% with the evaluation function

% \begin{lstlisting}
% churchEval :: Church r -> r
% churchEval (ChurchPair x y f) = f x y
% churchEval (ChurchLeft x f g) = f x
% churchEval (ChurchRight y f g) = g y
% \end{lstlisting}

% However, we only want to keep track of the functions used for elimination. By moving
% the values actually being contained in the products and coproducts out to the
% evaluation function, we arrive at

% \begin{lstlisting}
% data Church' t r where
%   ChurchPair'
%     :: (a -> b -> r) ->
%        Church' (a, b) r

%   ChurchEither'
%     :: (a -> r) -> (b -> r) ->
%        Church' (Either a b) r

% churchEval' :: Church' t r -> t -> r
% churchEval' (ChurchPair' f) (x, y) = f x y
% churchEval' (ChurchEither' f g) e =
%   case e of
%     Left  x -> f x
%     Right y -> g y
% \end{lstlisting}

% The \verb|Church'| type forms a large part of the foundation of the pattern
% matching aspect of the \verb|E| type. A rough correspondence between the two
% is given in the table below.\\

% {\small
% \noindent\begin{tabular}{|p{3.4cm}|p{4.4cm}|}
%   \hline
%   \ttt{Church' (a, b) r} & \ttt{E (SumMatch (a, b) r} \\
%   \hline
%   \ttt{Church' (Either a b) r} & \ttt{E (ProdMatch (Either a b) r)}\\
%   \hline
% \end{tabular}
% }

@@

% \newpage

% \section{C Backend}
% \label{sec:CBackend}

% \begin{todont}
%   Should this section be expanded?
% \end{todont}

% The C backend makes use of a small EDSL for generating C code. See
% Appendix B for implementation details of this ESDL.

% The following \ttt{struct}s provide the C representation for Haskell expressions:

% \begin{lstlisting}
% typedef enum var_type_tag {
%   EXPR_INT
% , EXPR_PAIR
% ...
% , EXPR_UNIT
% } var_type_tag;

% typedef enum semantic_type_tag {
%   NO_SEMANTIC_TAG
% , EXPR_COMPLEX     // Complex numbers
% } semantic_type_tag;

% struct closure_t;

% typedef struct var_t {
%   var_type_tag tag;
%   semantic_type_tag semantic_tag;
%   void* value;
% } var_t;

% typedef struct closure_t {
%   var_t* fv_env;
%   var_t (*fn)(var_t, struct closure_t*);
% } closure_t;
% \end{lstlisting}

% The \ttt{semantic\_tag} allows type information preserved by the \ttt{Typeable} constraints
% in the EDSL to be brought over to the generated C code. In our implementation, this is
% used to distinguish between pairs of \ttt{Double}s and \ttt{Complex Double} values. In the
% case of a \ttt{Complex Double}, the \ttt{var\_t} would have a \ttt{tag} field with the value
% \ttt{EXPR\_PAIR} and a \ttt{semantic\_tag} field with the value \ttt{EXPR\_COMPLEX}.

% A C code generation function \ttt{genC} is implemented:

% \begin{lstlisting}
% genC :: E a -> String
% ...
% \end{lstlisting}

% To use this, for example, with the previously mentioned \ttt{isEven} implementation we
% can write:
% % \newpage
% \begin{lstlisting}
% isEvenE :: E (Int -> Bool)
% isEvenE =
%   externalize (
%     case x == 0 of
%       True -> True
%       False ->
%         case x == 1 of
%           True -> False
%           False -> isEvenE (x - 2))

% isEvenC :: String
% isEvenC = genC isEvenE
% \end{lstlisting}

@@

  \begin{wrapfigure}{R}{0.6\linewidth}

@@

\resizebox{0.9\linewidth}{!}{
